Pricing Patterns in Organic Markets

### 2024-12-04 paper/paper.qmd
Update paper
paper fixed grammar and prose

---
title: "Pricing Patterns in Organic Markets"
subtitle: " Comparative Analysis of Vendor Strategies and Seasonal Trends in Organic Product Sales"
author: 
  - Dingshuo Li
thanks: "Code and data are available at: [https://github.com/dawsonlll/Organic-Product-Pricing-Analysis.git](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format:
  pdf:
    toc: true
  html:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

The organic food market has witnessed a remarkable surge in demand, driven by increasing consumer awareness of health and environmental impacts. This demand has not only heightened the market value of organic products but has also introduced complexities into their pricing strategies across various retail platforms. Notably, organic food typically incurs higher production and certification costs compared to conventional food products, influencing its retail pricing and accessibility. This study aims to dissect these pricing strategies by examining how the prices of organic foods vary across different vendors and over successive months.

This paper explores the intricate dynamics of organic food pricing within a well-defined dataset that records prices from multiple vendors over a specified period. The analysis is structured to address a gap in existing research: the detailed quantification of how prices for organic products are structured and vary in a real-market scenario. Utilizing advanced Bayesian statistical methods, this research assesses the relationship between current and historical prices while considering the influence of temporal factors captured through monthly variations.

Estimand: The primary estimand of this study is the price elasticity of organic foods across different market vendors and months. Specifically, the analysis seeks to estimate how changes in the historical prices (old prices) influence the current pricing strategies, adjusting for variations across months and between vendors. This will allow us to quantify the direct and interactive effects of time (month) and vendor on the current prices of organic products, providing insights into the responsiveness of organic food prices to market dynamics and vendor-specific factors.

The findings reveal distinct pricing patterns that differ by vendor and demonstrate notable fluctuations corresponding to seasonal trends. These insights are crucial for consumers making informed purchasing decisions and for vendors strategizing to capture or expand market share in the competitive organic food sector. The importance of understanding these pricing mechanisms lies in its potential to influence marketing strategies, consumer spending, and ultimately, the economic sustainability of organic farming practices.

The paper is structured to provide a clear, logical flow from methodology to implications @sec-data outlines our dataset selection and cleaning, ensuring transparency. The @sec-model details the Bayesian model and its rationale, directly linked to our data preparation. The @sec-result displays our findings with visual aids, while the @sec-discussion interprets these findings in the context of broader market dynamics. Finally, @sec-limitation  highlights areas for further study, acknowledging any constraints encountered.

@@

### Data Source
The dataset under analysis was sourced from Jacob Filipp's Hammer Project @hammerData, which compiles extensive pricing data on organic food products sold across various retail vendors. This dataset is particularly valuable for its comprehensive coverage of both temporal and vendor-specific pricing strategies, offering a unique lens through which seasonal trends and competitive behaviors in the organic food market can be observed. The choice of this dataset was motivated by its granularity and the quality of its data, which includes multiple vendors and a wide temporal range, making it ideal for a nuanced analysis of pricing dynamics.

@@

The dataset used in this study comes from Jacob Filipp's Hammer Project, which provides detailed records of organic food pricing across retailers and months. This data collection effort captures an important economic and social phenomenon: fluctuating pricing strategies in the competitive organic food market. Prices are recorded monthly to reflect the dynamic nature of these market conditions, providing a comprehensive snapshot that facilitates analysis of trends and pricing behavior over time.

The primary step in capturing the phenomenon involves systematically collecting price data directly from retail sources. This data is gathered from grocery merchants. Each price tag observed represents a specific decision by a vendor regarding how much to charge for an organic product at a given time. So the important variables: current price which is the price of an organic product at the time of data collection. It is observed directly from the vendor. And old price which is the previously recorded price for the same product, which provides a historical comparison point. It helps in understanding how prices are adjusted over time.

Each recorded price becomes an entry in the dataset, associated with metadata that includes the vendor, the time of collection, and the product details. This structured approach ensures that each entry in the dataset is a reliable and accurate representation of the real-world pricing phenomena observed. These entries are then ready for analytical processes that can derive meaningful insights about organic food pricing dynamics.

## Data Cleanning
The analysis begins with the consolidation of raw data from two distinct sources: product transaction records and product details. These datasets were sourced from the Hammer Project, containing over 12 million rows of diverse product data, including non-organic items and a variety of vendors beyond the scope of this study. The transaction records raw_data include timestamps, vendor details, and product identifiers, while the product details product_data provide comprehensive information such as product names, brands, pricing history, and unit measurements.

To conduct analysis, the two datasets were merged using an inner join on the product_id, which serves as a unique identifier linking transaction records to specific product details. This joining process ensured that each transaction record was enriched with relevant product information, creating a comprehensive view of each sale's context.

Post-merger, the dataset still contained an extensive range of products and vendors, many of which were irrelevant to the focus of this study—organic food pricing dynamics among major vendors. To refine the dataset first did vendor selection. The dataset was filtered to include only transactions from major retailers known for their significant market presence and influence in organic food sales, specifically "Metro," "NoFrills," "Walmart," "Voila," "TandT," and "Galleria." This selection was based on the objective to analyze pricing strategies within the most competitive segments of the retail market. A further filtration was applied to isolate organic products. This was achieved by searching for the keyword "organic" within the product_name field, using a case-insensitive search to ensure comprehensive inclusion of all relevant items. This step significantly reduced the dataset size and increased its relevance to the study’s aims.

The current_price and old_price were parsed to ensure that they were represented as numerical values, free from any non-numeric characters that could interfere with quantitative analysis. This parsing was crucial for accurate statistical calculations and comparisons. The nowtime field, representing the timestamp of each transaction, was used to extract the month component. This transformation was essential for analyzing seasonal trends and month-to-month pricing variations. Lastly, entries with missing or zero values in current_price were removed to maintain the integrity of pricing analysis. Additionally, any remaining rows with incomplete data were dropped, ensuring that the dataset used in subsequent analyses was entirely robust and clean.

## Outcome variables
### Predictor variables
The old price variable, representing the product's price from the preceding month, serves as a critical metric for understanding the evolution of market trends and pricing strategies over time. The summary @tbl-old-price reveal a broad range of "Old Price" from a minimum of \$1.69 to a maximum of \$38.56, with a median price of \$5.49, suggesting a median market position. The mean, slightly higher at \$7.66, indicates a positive skew in the data, with the first quartile at \$3.99 and the third at $10.49, reflecting substantial variability in historical pricing. This variability underscores the diverse pricing strategies across different vendors and product types, and provides a foundation for analyzing how prices adjust in response to market dynamics, consumer demand, and possibly supply chain fluctuations.

```{r}
#| label: tbl-old-price
#| tbl-cap: Summary of old price
#| echo: false
#| warning: false
#| message: false
old_price_summary <- summary(organic_analysis_data$old_price)

@@

# Print the summary as a table
kable(old_price_summary_df, format = "markdown")
```
The vendor variable acts as a categorical identifier for retailers within the dataset, providing crucial insights into the diversity of pricing strategies employed across different retail environments. This variable is essential for comparing and contrasting market behaviours as it delineates the dataset into distinct groups based on the retailer. The distribution of data @tbl-Vendor shows that vendors varies significantly, with Metro being the most frequently occurring vendor with 1,981 entries, followed by Voila with 1,513 entries. Walmart, Galleria, TandT, and NoFrills appear with decreasing frequency, suggesting a wide range in the scale and possibly the geographic reach of these retailers. This spread indicates that Metro and Voila dominate this dataset, which could imply a greater influence or market share in the organic products sector compared to the other vendors. Understanding the pricing strategies of these vendors is key to analyzing market trends, as each vendor may react differently to market pressures, consumer preferences, and seasonal changes. Analyzing this variable can reveal patterns that are specific to certain types of retailers and provide valuable insights into the competitive landscape of the organic food market.
```{r}
#| label: tbl-Vendor
#| tbl-cap: Preview of Breakdown of Organic Product Listings by Vendor
#| echo: false
#| warning: false
#| message: false
vendor_summary <- organic_analysis_data %>%
  count(vendor) %>%
  arrange(desc(n))

@@

```
The "Month" variable in the dataset, recorded as a numerical value, serves as a crucial indicator for analyzing seasonal impacts on organic food pricing. It reflects when the data was collected, revealing essential patterns of how prices fluctuate through different months, likely influenced by consumer demand, supply variations, and promotional periods. The distribution of observations shown in @tbl-month, the peaks are in summer season suggests significant market activity during the summer months, potentially driven by an increased availability of fresh produce or higher consumer buying rates. In contrast, the notably lower entries in spring and winter indicate reduced activity, which could be due to offseason market slowness or less frequent data collection. This variable is integral to understanding how temporal factors influence pricing strategies, providing valuable insights for vendors, consumers, and analysts studying market dynamics in the organic food sector.
```{r}
#| label: tbl-month
#| tbl-cap: Preview of Breakdown of Organic Product Listings by Month
#| echo: false
#| warning: false
#| message: false
month_summary <- organic_analysis_data %>%
  count(month) %>%
  arrange(month)
kable(month_summary, format = "markdown", col.names = c("Month", "Count"))

@@

### Response Variable
The response variable current price represents the most recent price at which organic food products were sold across various vendors. The summary statistics @tbl-current-price for this variable provide a snapshot of its distribution within the dataset. The minimum price recorded is 0.97, indicating a lower-end pricing point for some products, while the maximum price peaks at 29.99, reflecting premium offerings within the organic sector. The first quartile is 3.22, suggesting that at least 25% of the products are priced at or below this level, which could be indicative of basic or smaller-sized organic goods. The median price of 4.49 and the mean of 5.69 suggest a central tendency towards moderately priced items, with the mean being slightly higher due to the influence of more expensive products pulling the average up. The third quartile at 7.49 shows that 75% of the products fall below this price point, highlighting a pricing structure that might cater more to a mid-range consumer segment. This comprehensive look at current price helps in understanding the economic landscape of organic food pricing and its variability, which is crucial for analyzing market strategies and consumer behaviour.
```{r}
#| label: tbl-current-price
#| tbl-cap: Preview of Summary of Current Price
#| echo: false
#| warning: false
#| message: false
current_price_summary <- summary(organic_analysis_data$current_price)

@@

# Model
## Model Overview
In our study, we are particularly interested in understanding how the current price of organic products is influenced by several key factors: the price of the product in the previous month (old price), the vendor from which the product is purchased, and the time of year (month). To systematically examine these relationships, we developed a Bayesian regression model. This model incorporates these variables to elucidate their impacts on current pricing strategies. By leveraging historical pricing data alongside vendor-specific information and temporal dynamics, our model aims to uncover meaningful patterns that can inform both theoretical understanding and practical applications in market pricing strategies.

@@

```
We conducted the Bayesian analysis in R [@citeR], utilizing the `rstanarm` package [@rstanarm]. To maintain consistency and reliability in our modeling approach, we employed the default priors provided by `rstanarm`. These priors are well-suited for a wide range of data types and ensure robustness in the inference process, particularly in complex models such as ours that involve multiple predictors and hierarchical structures.

Normal priors with a mean of 0 and a standard deviation of 2.5 are assigned to the coefficients and the intercept. These priors were chosen to impose a slight regularization effect, reducing the risk of overfitting by moderating the influence of extreme values or outliers in the data.

@@

Unlike frequentist regression, Bayesian regression provides a probabilistic approach to inference, offering a full posterior distribution of the parameters, which helps in understanding the uncertainty around the estimated effects. This model is robust to various types of data and does not strictly require the normality assumption for residuals that a classical linear regression would. This is particularly beneficial given the often skewed or heterogeneous nature of pricing data. Each predictor's influence is quantified by respective coefficients, adjusted for vendor-specific variations and temporal trends. The use of normal priors for these coefficients is justified by the need to regularize estimates, thus preventing overfitting and ensuring stable predictions even with potentially collinear or sparse data.

### Comparison with Alternative Models
Although a standard multiple linear regression could model continuous outcomes effectively, it lacks the ability to incorporate prior distributions and assess uncertainty comprehensively. odels such as logistic regression or Poisson regression were considered; however, these models are more suited to categorical or count data, respectively. Given that our outcome variable, current price is continuous, a linear approach is more appropriate. These were also evaluated, especially for capturing any non-linear dynamics between the predictors and the current price. While potentially offering a closer fit to certain complex patterns, the increased model complexity could hinder interpretation and require more extensive data to validate effectively.

@@

# Results

Our results are summarized in @tbl-modelresults.

```{r tbl.pos='H'}
#| label: tbl-modelresults
#| tbl-cap: Summary of Bayesian Model
#| echo: false
#| eval: true
#| warning: false
#| message: false
modelsummary::modelsummary(
  list(
    "Bayesian_model" = bayesian_price_model
  ),
  statistic = "mad",
  fmt = 2
)

```
The Bayesian regression model's outputs provide insightful revelations about the factors influencing the current prices of organic products. These findings are encapsulated in Table 5, which succinctly presents the estimated coefficients and their respective standard errors. This quantitative output, complemented by the visual representation in the corresponding plot, offers a clear perspective on the relative impacts of each predictor.

The intercept, estimated at 3.53, represents the average base price of organic products when all other predictor variables are held at zero. This baseline offers a benchmark against which the effects of other variables are measured.The coefficient for month, which is -0.42, suggests a monthly decrease in prices. This trend could be indicative of seasonal adjustments, where prices may drop during certain times of the year due to increased supply from harvest periods or reduced consumer demand. The negative value emphasizes the importance of considering temporal factors in pricing strategies. The positive coefficient for old_price (0.62) underscores a significant reliance on historical pricing data. This relationship indicates that prices are likely to increase if they were higher in the preceding period, reflecting perhaps market confidence or ongoing trends in consumer willingness to pay.

The vendor coefficients reveal distinct pricing strategies among different retailer. vendorVoila shows a notably higher positive coefficient of 1.1), suggesting that this vendor typically prices organic products higher than others. This could reflect a premium branding strategy or a targeted demographic willing to pay more for perceived quality. In contrast, vendorNoFrills and vendorTandT have coefficients of 0.09 and -0.02, respectively, indicating minimal deviation from the baseline prices, possibly due to a competitive pricing strategy aimed at attracting cost-conscious consumers.

```{r fig.pos='H'}
#| label: fig-vendor
#| fig-cap: Outliers and trends in Pricing Between Different Vendors
#| echo: false
#| eval: true
#| warning: false
#| message: false

organic_analysis_data$price_change = organic_analysis_data$current_price - organic_analysis_data$old_price

# Scatter plot of Price Change vs. Month, color-coded by Vendor
ggplot(organic_analysis_data, aes(x = month, y = price_change, color = vendor)) +
  geom_point(alpha = 0.6, size = 3) +
  geom_smooth(se = FALSE) +
  labs(title = "Monthly Price Change by Vendor",
       x = "Month",
       y = "Price Change ($)",
       color = "Vendor") +
  theme_minimal()

```
The @fig-vendor illustrates the changes in pricing for organic products offered by six vendors over three key months: March, June, and September. It reveals that most vendors have adjusted their prices moderately, with the majority of changes clustering a little below the $5 mark. This pattern suggests a trend where vendors generally reduce their prices slightly, potentially reflecting seasonal promotions or responses to market dynamics. Notably, some vendors like Metro exhibit more dramatic price reductions, indicating aggressive pricing strategies or perhaps an adjustment to overstock situations. This subtle yet consistent price reduction across most vendors could imply a market-wide strategy to boost sales volumes by making organic products more affordable to a broader customer base during these months.

```{r fig.pos='H'}
#| label: fig-price
#| fig-cap: Price Trends in Different Vendor Over Months
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(organic_analysis_data, aes(x = month, y = current_price, color = vendor)) +
  geom_line(aes(group = interaction(vendor, month)), linewidth = 1) +  # Updated line
  geom_line(aes(y = old_price), linetype = "dashed") +
  facet_wrap(~vendor) +
  labs(title = "Price Trends by Vendor and Month",
       x = "Month",
       y = "Price ($)",
       color = "Vendor") +
  theme_minimal()
```
In the @fig-price, the pricing trends of presumably organic food items across six different vendors are displayed over months. The graph is segmented into six panels, each corresponding to a vendor: Galleria, Metro, NoFrills, TandT, Voila, and Walmart, allowing for a direct comparison of price fluctuations across different time points within the year. Each vendor's pricing data is presented with a unique colour code, maintaining consistency across the panels, which aids in quick visual comparison and analysis.

The analysis of the graph reveals distinct trends and behaviors among the vendors. Galleria shows a minor incremental trend in pricing, suggesting a stable market position. In contrast, Metro displays notable volatility, with significant price spikes evident in June. NoFrills exhibits very limited data variability in the graph, which could be indicative of several market behaviors or strategies. This minimal fluctuation in pricing might suggest that NoFrills does not have a significant inventory of organic food products, perhaps due to a strategic focus on other product lines or a customer base that is less inclined towards organic options.  TandT's prices slightly decline over the observed months, potentially reflecting seasonal promotions or inventory adjustments. Voila exhibits the most variability, with prices peaking significantly in summer seasons, which may reflect dynamic pricing strategies in response to demand changes or supply issues. Walmart, similar to Voila, shows price fluctuations with an increasing trend towards variability as the year progresses.

```{r fig.pos='H'}
#| label: fig-trends
#| fig-cap: Difference between Old Price and Current Price of Organic product
#| eval: true
#| echo: false
#| warning: false
#| message: false
ggplot(organic_analysis_data, aes(x = month)) +
  geom_smooth(aes(y = current_price, colour = "Current Price"), se = FALSE) +
  geom_smooth(aes(y = old_price, colour = "Old Price"), se = FALSE) +
  labs(title = "Current and Old Prices Over Time",
       x = "Month",
       y = "Price ($)",
       color = "Price Type") +
  scale_color_manual(values = c("Current Price" = "blue", "Old Price" = "red")) +
  theme_minimal()

>>>>

---
title: "Pricing Patterns in Organic Markets"
subtitle: " Comparative Analysis of Vendor Strategies and Seasonal Trends in Organic Product Sales"
author: 
  - Dingshuo Li
thanks: "Code and data are available at: [https://github.com/dawsonlll/Organic-Product-Pricing-Analysis.git](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "This paper investigates pricing patterns in the organic food market by analyzing how vendor strategies and seasonal trends influence product prices. Using a dataset of organic food prices across multiple vendors and months, we employed a Bayesian regression model to understand the interplay between historical prices, vendor-specific strategies, and seasonal factors. The analysis revealed significant seasonal price fluctuations and demonstrated that vendors rely heavily on historical prices to inform current pricing strategies. These findings provide valuable ideas into consumer behaviour, vendor decision-making, and the economic dynamics of organic markets, highlighting the broader implications for market competition and sustainable agricultural practices."
format:
  pdf:
    toc: true
  html:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

The organic food market has experienced a significant increase in demand, driven by growing consumer awareness of health and environmental impacts. This demand has not only raised the market value of organic products but has also introduced challenges to their pricing strategies across various retail platforms. Organic foods generally incur higher production and certification costs compared to conventional products, which affects their retail pricing and accessibility. This study examines these pricing strategies by analyzing how organic food prices vary across vendors and over time.

This paper investigates the dynamics of organic food pricing using a dataset that records prices from multiple vendors over a specific period. The analysis addresses the need for detailed quantification of how prices for organic products are structured and vary in real-market conditions. By applying Bayesian statistical methods, this research evaluates the relationship between current and historical prices while considering the effect of temporal factors, such as monthly variations.

Estimand: The primary focus of this study is the price elasticity of organic foods across different vendors and time periods. Specifically, the analysis estimates how changes in old prices influence current pricing strategies, accounting for differences across months and vendors. This approach allows for the measurement of the direct and interactive effects of month and vendor on the current prices of organic products, offering a deeper understanding of the responsiveness of organic food prices to market conditions and vendor-specific factors.

The findings highlight distinct pricing patterns among vendors and show significant fluctuations aligned with seasonal trends. These results are relevant for consumers making informed purchasing decisions and for vendors optimizing strategies to maintain or grow their market share in the competitive organic food sector. Understanding these pricing mechanisms has implications for marketing, consumer behavior, and the economic sustainability of organic farming practices.

The paper is structured to provide a clear, logical flow from methodology to implications @sec-data outlines our dataset selection and cleaning, ensuring transparency. The @sec-model details the Bayesian model and its justifications. The @sec-result displays our findings with visual graphs, while the @sec-discussion interprets these findings in the context of market dynamics. As well as highlights areas for further study, and the limitation of study.

@@

### Data Source
The dataset analyzed in this study was sourced from Jacob Filipp's Hammer Project @hammerData, which compiles detailed pricing data on organic food products across various retail vendors. Its extensive coverage of vendor-specific and temporal pricing strategies makes it particularly suitable for observing seasonal trends and competitive behaviors in the organic food market. The dataset’s granularity and quality, including data from multiple vendors over an extended period, provide a robust foundation for analyzing pricing dynamics.

@@

The dataset analyzed in this study provides detailed records of organic food prices across vendors and months, capturing the dynamic pricing strategies within the competitive organic food market. Monthly price recordings reflect changing market conditions and facilitate the analysis of trends and pricing behavior over time.

The data collection involved systematically recording prices directly from grocery merchants. Each price entry represents a vendor’s decision on how much to charge for an organic product at a given time. Two key variables are examined: the current price, observed directly at the time of collection, and the old price, the previously recorded price for the same product. The old price serves as a historical reference, aiding in the analysis of price adjustments over time.

Each entry in the dataset includes metadata detailing the vendor, collection date, and product information, ensuring accuracy and reliability. This structured approach supports robust analysis and allows for meaningful conclusions about organic food pricing dynamics.

## Data Cleanning
The analysis began with the consolidation of raw data from two sources: product records and product details. These datasets contained over 13 million rows of diverse product data, including non-organic items and vendors outside the scope of this study. The product records raw_data included timestamps, vendor details, and product identifiers, while the product details dataset product_data provided information such as product names, brands, pricing history, and unit measurements.

To prepare the data, the two datasets were merged using an inner join on the product_id, a unique identifier linking transaction records to specific product details. This process enriched each transaction record with relevant product information, creating a detailed view of each product.

After merging, the dataset still included a wide range of products and vendors, many of which were irrelevant to the study’s focus on organic food pricing among major vendors. The first step in refining the dataset was vendor selection. Transactions were filtered to include only those from major vendors with a significant presence in organic food sales: "Metro," "NoFrills," "Walmart," "Voila," "TandT," and "Galleria." This selection focused the analysis on pricing strategies within competitive segments of the retail market.

The current price and old price were parsed to ensure that they were represented as numerical values, free from any non-numeric characters that could interfere with quantitative analysis. This parsing was important for accurate statistical calculations and comparisons. The nowtime field, representing the timestamp of each transaction, was used to extract the month component. This transformation was essential for analyzing seasonal trends and month-to-month pricing variations. Lastly, entries with missing or zero values in current price were removed to maintain the integrity of pricing analysis. Additionally, any remaining rows with incomplete data were dropped, ensuring that the dataset used in subsequent analyses was entirely robust and clean.

## Outcome variables
### Predictor variables
The old price variable, representing the product's price from the preceding month, serves as a essential metric for understanding the evolution of market trends and pricing strategies over time. The summary @tbl-old-price reveal a broad range of "Old Price" from a minimum of \$1.69 to a maximum of \$38.56, with a median price of \$5.49, suggesting a median market position. The mean, slightly higher at \$7.66, indicates a positive skew in the data, with the first quartile at \$3.99 and the third at $10.49, reflecting substantial variability in historical pricing. This variability underscores the diverse pricing strategies across different vendors and product types, and provides a foundation for analyzing how prices adjust in response to market dynamics, consumer demand, and possibly supply chain fluctuations.

```{r}
#| label: tbl-old-price
#| tbl-cap: Overview of Historical Price Trends
#| echo: false
#| warning: false
#| message: false
old_price_summary <- summary(organic_analysis_data$old_price)

@@

# Print the summary as a table
kable(old_price_summary_df, format = "markdown")
```
The vendor variable serves as a categorical identifier for vendors in the dataset, enabling the analysis of pricing strategies across different retail environments. It is essential for comparing and contrasting market behaviors, as it divides the dataset into distinct groups based on the retailer. The distribution of data @tbl-Vendor shows significant variation among vendors, with Metro being the most frequently represented, with 1,981 entries, followed by Voila with 1,513 entries. Walmart, Galleria, TandT, and NoFrills appear less frequently, reflecting differences in the scale and possibly the geographic reach of these vendors

This distribution suggests that Metro and Voila dominate the dataset, potentially indicating a larger influence or market share in the organic products sector compared to other vendors. Understanding the pricing strategies of these vendors is important for analyzing market trends, as each vendor may respond differently to market pressures, consumer preferences, and seasonal changes. Analyzing this variable helps identify patterns specific to certain types of vendors and provides a better understanding of the competitive dynamics of the organic food market.
```{r}
#| label: tbl-Vendor
#| tbl-cap: Distribution of Organic Product Listings Across Vendors
#| echo: false
#| warning: false
#| message: false
vendor_summary <- organic_analysis_data %>%
  count(vendor) %>%
  arrange(desc(n))

@@

```
The "Month" variable in the dataset, recorded as a numerical value, serves as a important indicator for analyzing seasonal impacts on organic food pricing. It reflects when the data was collected, revealing essential patterns of how prices fluctuate through different months, likely influenced by consumer demand, supply variations, and promotional periods. The distribution of observations shown in @tbl-month, the peaks are in summer season suggests significant market activity during the summer months, potentially driven by an increased availability of fresh produce or higher consumer buying rates. In contrast, the notably lower entries in spring and winter indicate reduced activity, which could be due to offseason market slowness or less frequent data collection. This variable is integral to understanding how temporal factors influence pricing strategies, providing valuable strategy for vendors, consumers, and analysts studying market dynamics in the organic food sector.
```{r}
#| label: tbl-month
#| tbl-cap: Monthly Distribution of Organic Product Listings
#| echo: false
#| warning: false
#| message: false
month_summary <- organic_analysis_data %>%
  count(month) %>%
  arrange(month)
kable(month_summary, format = "markdown", col.names = c("Month", "Count"))

@@

### Response Variable
The response variable current price represents the most recent price at which organic food products were sold across various vendors. The summary statistics @tbl-current-price for this variable provide a snapshot of its distribution within the dataset. The minimum price recorded is 0.97, indicating a lower-end pricing point for some products, while the maximum price peaks at 29.99, reflecting premium offerings within the organic sector. The first quartile is 3.22, suggesting that at least 25% of the products are priced at or below this level, which could be indicative of basic or smaller-sized organic goods. The median price of 4.49 and the mean of 5.69 suggest a central tendency towards moderately priced items, with the mean being slightly higher due to the influence of more expensive products pulling the average up. The third quartile at 7.49 shows that 75% of the products fall below this price point, highlighting a pricing structure that might cater more to a mid-range consumer segment. This look at current price helps in understanding the economic landscape of organic food pricing and its variability, which is important for analyzing market strategies and consumer behaviour.
```{r}
#| label: tbl-current-price
#| tbl-cap: Summary of Current Price Distribution
#| echo: false
#| warning: false
#| message: false
current_price_summary <- summary(organic_analysis_data$current_price)

@@

# Model {#sec-model}
## Model Overview
In our study, we are particularly interested in understanding how the current price of organic products is influenced by several key factors: the price of the product in the previous month, the vendor from which the product is purchased, and the time of year. To systematically examine these relationships, we developed a Bayesian regression model. This model incorporates these variables to elucidate their impacts on current pricing strategies. By leveraging historical pricing data alongside vendor-specific information and temporal dynamics, our model aims to uncover meaningful patterns that can inform both theoretical understanding and practical applications in market pricing strategies.

@@

```
We conducted the Bayesian analysis in R [@citeR], utilizing the `rstanarm` package [@rstanarm]. To maintain consistency and reliability in our modeling approach, we employed the default priors. These priors are well-suited for a wide range of data types and ensure robustness in the inference process, particularly in complex models such as ours that involve multiple predictors and hierarchical structures. Normal priors with a mean of 0 and a standard deviation of 2.5 are assigned to the coefficients and the intercept. These priors were chosen to impose a slight regularization effect, reducing the risk of overfitting by moderating the influence of extreme values or outliers in the data.

@@

Unlike other regression, Bayesian regression provides a probabilistic approach to inference, offering a full posterior distribution of the parameters, which helps in understanding the uncertainty around the estimated effects. This model is robust to various types of data and does not strictly require the normality assumption for residuals that a classical linear regression would. This is particularly beneficial given the often skewed or heterogeneous nature of pricing data. Each predictor's influence is quantified by respective coefficients, adjusted for vendor-specific variations and temporal trends. The use of normal priors for these coefficients is justified by the need to regularize estimates, thus preventing overfitting and ensuring stable predictions even with potentially collinear or sparse data.

### Comparison with Alternative Models
Although a standard multiple linear regression could model continuous outcomes effectively, it lacks the ability to incorporate prior distributions and assess uncertainty. odels such as logistic regression or Poisson regression were considered; however, these models are more suited to categorical or count data, respectively. Given that our outcome variable, current price is continuous, a linear approach is more appropriate. These were also evaluated, especially for capturing any non-linear dynamics between the predictors and the current price. While potentially offering a closer fit to certain complex patterns, the increased model complexity could hinder interpretation and require more extensive data to validate effectively.

@@

# Results {#sec-result}
```{r tbl.pos='H'}
#| label: tbl-modelresults
#| tbl-cap: Overview of Bayesian Model Analysis
#| echo: false
#| eval: true
#| warning: false
#| message: false
modelsummary::modelsummary(
  list(
    "Bayesian_model" = bayesian_price_model
  ),
  statistic = "mad",
  fmt = 2
)

```
Our results are summarized in @tbl-modelresults. The Bayesian regression model's outputs provide implications about the factors influencing the current prices of organic products. These findings are encapsulated in Table 5, which succinctly presents the estimated coefficients and their respective standard errors. This quantitative output, complemented by the visual representation in the corresponding plot, provides a clear perspective on the relative impacts of each predictor.

The intercept, estimated at 3.53, represents the average base price of organic products when all other predictor variables are held at zero. This baseline provide a benchmark against which the effects of other variables are measured.The coefficient for month, which is -0.42, suggests a monthly decrease in prices. This trend could be indicative of seasonal adjustments, where prices may drop during certain times of the year due to increased supply from harvest periods or reduced consumer demand. The negative value emphasizes the importance of considering temporal factors in pricing strategies. The positive coefficient for old_price (0.62) underscores a significant reliance on historical pricing data. This relationship indicates that prices are likely to increase if they were higher in the preceding period, reflecting perhaps market confidence or ongoing trends in consumer willingness to pay.

The vendor coefficients reveal distinct pricing strategies among different retailer. vendor Voila shows a notably higher positive coefficient of 1.1, suggesting that this vendor typically prices organic products higher than others. This could reflect a premium branding strategy or a targeted demographic willing to pay more for perceived quality. In contrast, NoFrills and TandT have coefficients of 0.09 and -0.02, respectively, indicating minimal deviation from the baseline prices, possibly due to a competitive pricing strategy aimed at attracting cost-conscious consumers.

```{r fig.pos='H'}
#| label: fig-vendor
#| fig-cap: Vendor-Specific Monthly Price Variations
#| echo: false
#| eval: true
#| warning: false
#| message: false

organic_analysis_data$price_change = organic_analysis_data$current_price - organic_analysis_data$old_price
ggplot(organic_analysis_data, aes(x = month, y = price_change, color = vendor)) +
  geom_point(alpha = 0.6, size = 3) +
  geom_smooth(se = FALSE) +
  labs(title = "Monthly Price Change by Vendor",
       x = "Month",
       y = "Price Change ($)",
       color = "Vendor") +
  theme_minimal()

```
The @fig-vendor illustrates the changes in pricing for organic products offered by six vendors over three key months: March, June, and September. It shows that most vendors have adjusted their prices moderately, with the majority of changes clustering a little below the $5 mark. This pattern suggests a trend where vendors generally reduce their prices slightly, potentially reflecting seasonal promotions or responses to market dynamics. Notably, some vendors like Metro exhibit more dramatic price reductions, indicating aggressive pricing strategies or perhaps an adjustment to overstock situations. This subtle yet consistent price reduction across most vendors could imply a market-wide strategy to boost sales volumes by making organic products more affordable to a broader customer base during these months.

```{r fig.pos='H'}
#| label: fig-price
#| fig-cap: Monthly Price Trends Across Vendors
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(organic_analysis_data, aes(x = month, y = current_price, color = vendor)) +
  geom_line(aes(group = interaction(vendor, month)), linewidth = 1) +  # Updated line
  geom_line(aes(y = old_price), linetype = "dashed") +
  facet_wrap(~vendor) +
  labs(title = "Price Trends by Vendor and Month",
       x = "Month",
       y = "Price ($)",
       color = "Vendor") +
  theme_minimal()
```
In the @fig-price, the pricing trends of presumably organic food items across six different vendors are displayed over months. The graph is segmented into six panels, each corresponding to a vendor: Galleria, Metro, NoFrills, TandT, Voila, and Walmart, allowing for a direct comparison of price fluctuations across different time points within the year. Each vendor's pricing data is presented with a unique colour code, maintaining consistency across the panels, which aids in quick visual comparison and analysis.

The analysis of the graph shows distinct trends and behaviors among the vendors. Galleria shows a minor incremental trend in pricing, suggesting a stable market position. In contrast, Metro displays notable volatility, with significant price spikes evident in June. NoFrills exhibits very limited data variability in the graph, which could be indicative of several market behaviors or strategies. This minimal fluctuation in pricing might suggest that NoFrills does not have a significant inventory of organic food products, perhaps due to a strategic focus on other product lines or a customer base that is less inclined towards organic options.  TandT's prices slightly decline over the observed months, potentially reflecting seasonal promotions or inventory adjustments. Voila exhibits the most variability, with prices peaking significantly in summer seasons, which may reflect dynamic pricing strategies in response to demand changes or supply issues. Walmart, similar to Voila, shows price fluctuations with an increasing trend towards variability as the year progresses.

```{r fig.pos='H'}
#| label: fig-trends
#| fig-cap: Comparison of Old and Current Prices for Organic Products
#| eval: true
#| echo: false
#| warning: false
#| message: false
ggplot(organic_analysis_data, aes(x = month)) +
  geom_smooth(aes(y = current_price, colour = "Current Price"), se = FALSE) +
  geom_smooth(aes(y = old_price, colour = "Old Price"), se = FALSE) +
  labs(title = "Current and Old Prices Over Time",
       x = "Month",
       y = "Price ($)",
       color = "Price Type") +
  scale_color_manual(values = c("Current Price" = "blue", "Old Price" = "red")) +
  theme_minimal()

```
The @fig-trends delineates the evolution of pricing data under two categories: "Current Price" and "Old Price," over the months. This graphical representation enables an comparison of the pricing fluctuations throughout the year for a particular product or set of products. The current price, represented in blue, shows a downward trajectory over the year, falling from about 8 dollars to approximately 4 dollars. Conversely, the old price, depicted in red, demonstrates a similar decline from March to June but maintains a level higher than the current price during this period. Notably, there is a sharp dip in September where it plunges to below 6 dollars before slightly recovering. This divergence between the current and old prices from June to September, where the current price escalates substantially while the old price sharply declines, may suggest responses to external market forces such as supply chain disruptions, variations in seasonal demand, or changes in promotional strategies that impact pricing structures. Particularly, the pronounced drop in the old price during September could indicate strategies such as clearance sales or pricing adjustments intended to clear existing inventory or adapt to waning demand.

This overall trend where the current prices fall below the old prices could indicate a reduction in the market price of organic food. Several factors could contribute to this phenomenon, such as increased supply of organic products, enhanced efficiency in organic farming techniques, or perhaps a decrease in consumer demand. Moreover, such a price drop could also be a strategic response by vendors to attract more customers or to remain competitive in the market, especially if other factors like seasonal variations or promotional efforts play a role.

# Discussion {#sec-discussion}
